Choosing Color Palettes for Statistical Graphics

### 2006-10-16 trunk/vcd/paper/Color/draft.Rnw
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/vcd@591 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

For choosing color palettes it is helpful to have an idea how
human color vision evolved. It has been hypothesized that it developed in three
distinct stages: 1.~perception of \textit{light/dark} contrasts (monochrome only),
2.~\textit{yellow/blue} contrasts (usually associated with our notion of warm/cold colors),
3.~\textit{green/red} contrasts (helpful for assessing the ripeness of fruit). See 
\cite{vcd:Ihaka:2003} for more details and references.

Due to these three color axes, colors are typically described as
locations in 3-dimensional spaces. However, human perception of color
does not correspond to the physiological axes above, but rather to polar
coordinates in the color plane (yellow/blue vs.~green/red) plus a third
light/dark axis. Thus, perceptually-based color spaces try to capture the
perceptual axes
\begin{enumerate}
  \item \textbf{hue} (dominant wavelength)
  \item \textbf{chroma} (colorfulness, intensity of color as compared to gray)
  \item \textbf{luminance} (brightness, amount of gray)
\end{enumerate}

A popular implementation of such a color space, available in many graphics and
statistics software packages, are \textit{HSV} (hue, saturation, value) colors.
They are a simple transformation of \textit{RGB} (red, green, blue) colors
and are defined by a triplet $(H, S, V)$ with $H \in [0, 360]$ and $S, V \in [0, 100]$.
HSV space has the shape of a single regular cone (often inflated to a regular
cylinder). Vertical sections through this space are shown in the upper panel
of Figure~\ref{fig:hsv-hcl}, depicting hue and saturation given different value levels.
Although simple to specify and easily available in many computing environments,
HSV colors have a fundamental drawback: its three dimensions map to the three
dimensions of human color perception very poorly. The three dimensions are
confounded which is most easily seen when converting the vertical sections to 
gray scale images in Figure~\ref{fig:hsv-hcl-gray}. Clearly, the brightness of
colors is not uniform over hues and saturations (given value)---therefore,
HSV colors are often not considered to be perceptually based.

To overcome these drawbacks, various color spaces have been suggested
that properly map to the perception dimensions, the most prominent of which are
the CIELUV and CIELAB spaces developed by the \cite{vcd:CIE:2004}. \cite{vcd:Ihaka:2003}
argues that CIELUV colors are typically preferred for use with emissive technologies
such as computer screens which makes them an obvious candidate for implementation
in statistical software packages. By taking polar coordinates in the UV plane of
CIELUV, \textit{HCL} (hue, chroma, luminance) colors are obtained, defined by
a triplet $(H, C, L)$ with $H \in [0, 360]$ and $C, L \in [0, 100]$. HCL space
has the shape of a distorted double cone: the admissable chroma and luminance values
depend on the hue chosen. The lower panel of Figure~\ref{fig:hsv-hcl} shows
vertical sections through this space: each of the resulting hue/chroma planes (given
luminance) is now properly balanced towards the same gray (going from black to white
with increasing luminance) which becomes obvious when converting the colors to
a gray scale is in Figure~\ref{fig:hsv-hcl-gray}. This balancing of HCL colors
gives us the opportunity to conveniently choose color palettes which code categorical
and/or numerical information by translating it to paths along the three perceptual axes.
However, some care is required for dealing with the irregular shape of the HCL space
which will be addressed in the following sections.

>>>>

For choosing color palettes it is helpful to have an idea how
human color vision evolved. It has been hypothesized that it developed in three
distinct stages: 1.~perception of \emph{light/dark} contrasts (monochrome only),
2.~\emph{yellow/blue} contrasts (usually associated with our notion of warm/cold colors),
3.~\emph{green/red} contrasts (helpful for assessing the ripeness of fruit). See 
\cite{vcd:Ihaka:2003} for more details and references.

Due to these three color axes, colors are typically described as
locations in 3-dimensional spaces. However, human perception of color
does not correspond to the physiological axes above, but rather to polar
coordinates in the color plane (yellow/blue vs.~green/red) plus a third
light/dark axis. Thus, perceptually-based color spaces try to capture the
perceptual axes
\begin{enumerate}
  \item \emph{hue} (dominant wavelength)
  \item \emph{chroma} (colorfulness, intensity of color as compared to gray)
  \item \emph{luminance} (brightness, amount of gray)
\end{enumerate}

A popular implementation of such a color space, available in many graphics and
statistics software packages, are \emph{HSV} (hue, saturation, value) colors.
They are a simple transformation of \emph{RGB} (red, green, blue) colors
and are defined by a triplet $(H, S, V)$ with $H \in [0, 360]$ and $S, V \in [0, 100]$.
HSV space has the shape of a single regular cone (often inflated to a regular
cylinder). Vertical sections through this space are shown in the upper panel
of Figure~\ref{fig:hsv-hcl}, depicting hue and saturation given different value levels.
Although simple to specify and easily available in many computing environments,
HSV colors have a fundamental drawback: its three dimensions map to the three
dimensions of human color perception very poorly. The three dimensions are
confounded which is most easily seen when converting the vertical sections to 
gray scale images in Figure~\ref{fig:hsv-hcl-gray}. Clearly, the brightness of
colors is not uniform over hues and saturations (given value)---therefore,
HSV colors are often not considered to be perceptually based.

To overcome these drawbacks, various color spaces have been suggested
that properly map to the perception dimensions, the most prominent of which are
the CIELUV and CIELAB spaces developed by the \cite{vcd:CIE:2004}. \cite{vcd:Ihaka:2003}
argues that CIELUV colors are typically preferred for use with emissive technologies
such as computer screens which makes them an obvious candidate for implementation
in statistical software packages. By taking polar coordinates in the UV plane of
CIELUV, \emph{HCL} (hue, chroma, luminance) colors are obtained, defined by
a triplet $(H, C, L)$ with $H \in [0, 360]$ and $C, L \in [0, 100]$. HCL space
has the shape of a distorted double cone: the admissable chroma and luminance values
depend on the hue chosen. The lower panel of Figure~\ref{fig:hsv-hcl} shows
vertical sections through this space: each of the resulting hue/chroma planes (given
luminance) is now properly balanced towards the same gray (going from black to white
with increasing luminance) which becomes obvious when converting the colors to
a gray scale is in Figure~\ref{fig:hsv-hcl-gray}. This balancing of HCL colors
gives us the opportunity to conveniently choose color palettes which code categorical
and/or numerical information by translating it to paths along the three perceptual axes.
However, some care is required for dealing with the irregular shape of the HCL space
which will be addressed in the following sections.
